Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Chinook Salmon Populations

### 2022-01-31 analysis/paper/paper.Rmd
incorporated TB discussion edits

The mean population size of Northern Pikeminnow in the Deadwater Slough during the fall season was estimated to be `r prettyNum(unique(round(impact_df$N[impact_df$season == "Fall"])), big.mark = ",")` and during the spring 2021 to be `r prettyNum(unique(round(impact_df$N[impact_df$season == "Spring"])), big.mark = ",")`. Those estimates translate to linear densities of `r prettyNum(round(unique(impact_df$N[impact_df$season == "Fall"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(unique(impact_df$N[impact_df$season == "Spring"]) / ddw_rkms), big.mark = ",")` 
<!-- need units  -->
which is `r round(unique(impact_df$N[impact_df$season == "Fall"]) / ddw_rkms / john_day_dens)` and `r round(unique(impact_df$N[impact_df$season == "Spring"]) / ddw_rkms / john_day_dens)` times greater, respectively, than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately `r as.character(english(round(unique(impact_df$N[impact_df$season == "Fall"]) / ddw_rkms / low_columbia_fish_km)))` and `r as.character(english(round(unique(impact_df$N[impact_df$season == "Spring"]) / ddw_rkms / low_columbia_fish_km)))` times greater than projections for the lower Columbia River [@Beamesderfer1996]. Although estimated densities for Deadwater Slough are high given the relative size (volume) of the Salmon River at the slough relative to the Columbia River, high CPUE combined with low recapture rates (one in every `r round(nrow(cmr_df) / sum(cmr_df$capture_type == "Recapture", na.rm = T))` captured fish were recaptures) suggest this area may support exceptionally high predator densities. Further, @Beamesderfer1991 acknowledge that their methods were unsuitable for sampling offshore, noting that water velocity, depth, irregular bottom contours, and barge traffic made sampling offshore ineffective. Conversely, at Deadwater Slough the maximum depth was approximately 10-15 m with a relatively homogenous bottom contour, making angling an effective method throughout the entire reach. Our estimates of Northern Pikeminnow areal density are also higher than those reported elsewhere in the Columbia River basin. We estimate the density of Northern Pikeminnow in Deadwater Slough, as fish per hectare, to be `r prettyNum(round(unique(impact_df$N[impact_df$season == "Fall"]) / ddw_hectares), big.mark = ",")` for fall and `r prettyNum(round(unique(impact_df$N[impact_df$season == "Spring"]) / ddw_hectares), big.mark = ",")` for spring. For comparison, the only other available estimates of areal density, that the authors are aware of, are 4.4 fish per hectare for John Day Reservoir [@Beamesderfer1991] and 15 fish per hectare in Lake Washington [@Bartoo1977]. Given the dearth of current data available in the Columbia River basin, it is unclear how Northern Pikeminnow abundance estimates in this study compare to elsewhere in the Columbia River Basin. @Widener2021 has identified a need for more current information on piscine predator populations in the Columbia River.

All mark-recapture abundance estimators suggest a large number of Northern Pikeminnow occupying Deadwater Slough. Several assumptions in our estimators may have influenced the magnitude of our results. First, all models assumed a closed population, however, it is possible that immigration and emigration occurred. Nevertheless, our estimators still provide unbiased estimates of abundance assuming the immigration and emigration rates are equal between marked and unmarked fish. Given the size of the sampling area, and the limited length of the sampling period (two weeks), that assumption is likely met. The assumption of a closed population can also affect results if mortality occurs for some marked fish released back to the population. During our field processing, we did not observe mortalities or injury post release; however, the potential for mortality amongst marked and released fish cannot be discounted. Next, our estimates assumed that the second (and subsequent) sample(s) are random samples of the population. Angling methods may have a size selection bias, thereby limiting our abundance estimates to a size threshold susceptible to angling, and possibly biasing them low. We also consider the validity of our assumption of equal catchability of individual fish between sampling events to have introduced potential bias in our results. Unequal catchability for individual fish is difficult to assess, especially if being caught once influences the likelihood of recapture. During our study, we recaptured two individual fish during multiple surveys indicating that marked fish were still susceptible to angling.<!--We are also comfortable with the assumptions of no loss of marks or no mistaken or missed marks given that marks were physical (hole punches in fins) and that each fish was examined on a measuring board to collect lengths.-->

We further recognize that marked and released fish may not have mixed thoroughly back into the population prior to recapture events. Although we attempted to release marked fish evenly throughout Deadwater Slough, we anecdotally observed anglers congregating in particular areas of the slough to socialize or exploit the "good" fishing locales. If marked fish were not recently released or mixed back into those areas that would result in marked fish being less susceptible to capture. Violations of the assumption of equal catchability between marked and unmarked fish may lead to an overestimate of abundance of Northern Pikeminnow in our study if marked fish have lower catchability. This bias will be proportional to the difference in catchability of marked and unmarked fish, so if marked fish are only 70% as likely to be caught as unmarked fish (due to behavior or spatial distribution) the true abundance is closer to 70% of our abundance estimate. We calculated this for the Schnabel estimator by assuming that only 70% of the fish marked each sample were returned to the populuation and available for recapture.

@@

The spring abundance estimate was smaller than fall abundance estimates. Our approach assumed equal capture probabilities between the fall and spring sample efforts. Yet, we suspect that the capture probability in the spring could be lower as higher spring flows reduce Northern Pikeminnow densities biasing our spring abundance estimate low. Therefore, we believe our spring abundance estimate to be conservative.
<!-- how do higher spring flows reduce pikeminnow densities? If pikeminnow densities are lower, then the total abundance SHOULD be lower. Makes more sense to me to say that higher flows directly lower capture probabilities. -->

@@

We performed gastric lavage on nearly all Northern Pikeminnow collected during this study. In addition to juvenile Chinook Salmon, species included Redside Shiner, Largescale Sucker *Catostomus macrocheilus*, sculpin *Cottus* spp., and Mountain Whitefish *Prosopium williamsoni*; although decomposition from digestion rendered many fish contents unidentifiable. While this approach provided useful information on prey species, we have concerns about using angling and gastric lavage to quantify diet composition in Northern Pikeminnow. First, previous studies have indicated that diets of angled fish differ from those collected by other methods (e.g., electrofishing) [@Hodgson1988], and further, angled fish are more likely to have empty stomachs [@Jurajda2016], perhaps because they are hungry and actively searching for food. Second, unlike other predators such as Smallmouth Bass and Walleye that are commonly sampled using gastric lavage [@Kamler2001], Northern Pikeminnow are cyprinids and therefore lack a true stomach. Previous research indicates that performing gastric lavage on cyprinid species can be ineffective at quantifying diet composition [@Hartleb1995], or alternately, requires a modification in methodology to flush food items out the vent of the fish [@Wasowicz1994]. Our intent, rather, was to simply document predation on juvenile Chinook Salmon in Deadwater Slough, and given the potential ineffectiveness of gastric lavage using angling and for Northern Pikeminnow, we consider our diet composition estimates to be conservative estimates i.e., we believe that a larger proportion of the Northern Pikeminnow diet is fish and juvenile Chinook Salmon than our gastric results suggest.

## Bioenergetics
<!--
OUR APPROACH:
1. g of fish consumed during DSR and NRR peak emigrations
2. Combine results from #1 for total g consumed by an individual NPM
3. Parse out the percentage of fish prey that is juvenile Chinook salmon
  - start w/ conservative scenarios of 10% and 25%
4. Divide total g consumed by 10.6 g, the average juvenile Chinook Salmon from Upper Salmon MPG, to estimate the # of juvenile Chinook Salmon consumed
5. Multiply results from #4 by a "stable" NPM population size estimate
6. Multiply results from #5 by SAR from McCann et al. 2019 for Upper Salmon MPG Chinook
-->

An average-size Northern Pikeminnow from Deadwater Slough is estimated to consume `r median(bio_g_summ$cumm_g[bio_g_summ$season == "Fall"])` g (range: `r min(bio_g_summ$cumm_g[bio_g_summ$season == "Fall"])` - `r max(bio_g_summ$cumm_g[bio_g_summ$season == "Fall"])` g) and `r median(bio_g_summ$cumm_g[bio_g_summ$season == "Spring"])` g (range: `r min(bio_g_summ$cumm_g[bio_g_summ$season == "Spring"])` - `r max(bio_g_summ$cumm_g[bio_g_summ$season == "Spring"])` g) of fish during fall and spring, respectively, to maintain its body weight. This equates to `r median(bio_g_summ$cumm_g[bio_g_summ$season == "Fall"]) + median(bio_g_summ$cumm_g[bio_g_summ$season == "Spring"])` g of fish consumed during the DSR and NRR peak emigrations, assuming the median value among diet scenarios (i.e., if 50% of their diet consists of fish). Consumption ranged from `r min(bio_g_summ$cumm_g[bio_g_summ$season == "Fall"]) + min(bio_g_summ$cumm_g[bio_g_summ$season == "Spring"])` for a diet comprised of 10% fish to `r max(bio_g_summ$cumm_g[bio_g_summ$season == "Fall"]) + max(bio_g_summ$cumm_g[bio_g_summ$season == "Spring"])` for a diet comprised of 90% fish.

Our approach assumed that all available prey consumed by Northern Pikeminnow have energy densities equal to juvenile Chinook Salmon. Other salmonid including juvenile steelhead and Sockeye Salmon are likely also available prey in Deadwater Slough, especially during the spring migration. Additionally, hatchery smolt releases (Chinook Salmon, steelhead, and Sockeye Salmon) are also prevalent in the mainstem Salmon River during the spring migration. Other studies have used a similar approach where the energy densities of different salmonids were generalized using an estimate from a single species [@Petersen1999].<!-- The following needs to be updated after we finalize some results --> However, our study is focused on potential impacts to natural-origin DSR and NRR Chinook Salmon, in which case, let's consider a thought exercise where DSR and NRR Chinook Salmon represent only 10% and 25% of total fish prey consumed by Northern Pikemnnow during these periods. Under these scenarios, the average Northern Pikeminnow would consume 8.95 g (range 4.52 – 10.05 g) if 10% of prey were natural-origin Chinook Salmon and 22.38 g (range 11.31 – 25.12) if 25%.

Our estimated consumption of Chinook Salmon juveniles by Northern Pikeminnow was calculated using a static predator population. If we assume an average juvenile Upper Salmon MPG Chinook Salmon of 10.6 g, this equates to 0.84 juvenile Chinook Salmon (range 0.43 – 0.94) under the 10% scenario and 2.11 juvenile Chinook Salmon (range 1.07 – 2.37) under the 25% scenario. Let’s now consider a stable Northern Pikeminnow population size of 10,000 (roughly the mean of fall and spring abundance estimates, divided by two to account for half of marked individuals not being available for recapture). In this case, the Northern Pikeminnow population in Deadwater Slough would be estimated to consume 8,400 (range 4,300 – 9,400) or 21,100 (range 10,700 – 23,700) under the 10% and 25% scenarios, respectively.

## Impacts to Adult Returns

<!-- Also update this after finalizing results -->
Our primary objective was to assess the impact of predation on juvenile Chinook Salmon in the Upper Salmon MPG. However, there is also interest in estimating the number of additional adult Chinook Salmon that might be expected to return to Lower Granite Dam if predation in the Deadwater Slough was reduced or eliminated. Our approach here was to multiply the total estimated juvenile Chinook Salmon consumed by an estimate of Granite-to-Granite smolt-to-adult survival rate (SAR) from @McCann2019 for Upper Salmon River MPG Chinook Salmon. Assuming an SAR of `r round(sar$medsar, 5)` (SE: `r prettyNum(round(sar$sdsar, 5))`) we'll consider the "adult equivalents" to our estimates of juveniles consumed. Under this scenario, the consumed juveniles would be equivalent to 51.6 adults (range 26.4 – 57.7) and 129.6 adults (range 65.7 – 145.6). However, we believe these 10% and 25% scenarios to be conservative, especially during the fall DSR emigration period. During spring, natural-origin steelhead and Sockeye Salmon are also actively emigrating from the Upper Salmon River and hatchery-origin releases of all three species (Chinook Salmon, steelhead, and Sockeye Salmon) are present in the river, and so 10% or 25% may be reasonable. However, during fall, DSR Chinook Salmon are likely the dominant prey species available to Chinook Salmon in Deadwater Slough as 1) no (or at least minimal) hatchery releases are present in the river and 2) fall emigrations of steelhead and Sockeye Salmon are less prominent than in Chinook Salmon. Note that this thought exercise only includes the Northern Pikeminnow population in Deadwater Slough even though they are present throughout the upper Salmon River, although likely at lower densities than in Deadwater Slough, and also ignores potential impacts to ESA-listed steelhead and Sockeye Salmon that are also a portion of the Northern Pikeminnow diet.
<!--
Additional thoughts here:
*	SAR results. Those results are GRA - GRA SARs, estimated from multiple years of PIT tag data. Seem pretty good to me.
* Maybe a plot of multiple scenarios
-->

@@

Although not formally assessed in this study, avian predators including Great Blue Herons *Ardea herodias* and Bald Eagles *Haliaeetus leucocephalus* are another potential source of mortality for juvenile Chinook Salmon and other ESA-listed salmonids in the Deadwater Slough. The Deadwater Slough is recognized as an important bird watching and nesting area due to the associated upstream braided channels and riparian and backwater habitats [(Deadwater Slough - Audubon Important Bird Areas)](https://www.audubon.org/important-bird-areas/deadwater-slough#). Several piscivorous bird species have been documented using Deadwater Slough including the Common *Mergus merganser* and Hooded *Lophodytes cucullatus* mergansers, the Great Blue Heron, the Double-crested Cormorant *Phalacrocorax auritus*, and the Belted Kingfisher *Megaceryle alcyon* [@eBird2021]. During the initial sampling effort in fall 2019, a two-person crew walked the entire reach including surrounding and upstream areas scanning for passive integrated transponder (PIT) tags that are implanted into thousands of natural- and hatchery-origin salmonids in the Upper Salmon River, annually. During that informal survey, 10 PIT tags were recovered near active bird nests and in an upstream anastomizing reach where herons and eagles are prevalent, suggesting that mortality may have been a result of avian predation. The PIT tag histories in [PTAGIS](https://www.ptagis.org/) indicate these tags were implanted into a combination of juvenile Chinook Salmon (3), Sockeye Salmon (4), and steelhead (3). Two of the Chinook Salmon and one steelhead were natural-origin, the remaining were of hatchery-origin. Avian predation contributes a major component of the total mortality for yearling Chinook Salmon in some locations in the lower Snake and Columbia rivers, particularly at hydroelectric dams and within reservoirs [@Evans2012; -@Evans2016]; however, we did not observe large colonies of piscivorous birds within the study area. Although we did observe Belted Kingfishers, mergansers, and Great Blue Herons at Deadwater Slough during field efforts; Great Blue Herons were especially prevalent in the anastomizing reach immediately upriver of the slackwater reach. Additionally, there is documentation of individual Double-crested Cormorants [@eBird2021] at the Deadwater Slough; however, the site is not within their breeding range, rather, it is part of a migration corridor. Given the current avian species known to occupy Deadwater Slough, it is unlikely that avian predation on juvenile salmonids is comparable to elsewhere in the Columbia River basin with large piscivorous bird colonies. Nevertheless, we hypothesize that the reservoir-like conditions and associated habitat features at the Deadwater Slough may increase the probability of avian predation on juvenile Chinook Salmon and other ESA-listed salmonids from the many piscivorous birds known to use the site. Future studies of predation would benefit from estimates from piscivorous avian predators.

@@

<!-- Again, some of the following numbers need to be updated after results are formalized -->
Recovery of Chinook Salmon in the Upper Salmon MPG is desired to support local fisheries and economies. At least five of the eight populations must meet criteria set forth by @McElhany2000 and @ICTRT2007 for the MPG to be considered viable and for recovery of the Snake River Evolutionary Significant Unit. Spawner abundance is perhaps the most important metric considered in determining a population's viability. We estimated that consumption of juvenile Chinook Salmon by Northern Pikeminnow in the Deadwater Slough potentially reduces natural-origin returns to upriver populations by 26.4 to 145.6 adults, annually. This is between 4.7% and 26.1% of the total mean adult escapement<!--of 558--> reported for the Upper Salmon River MPG in 2017 - 2019 [@Kinzer2020]. These estimates only take into account predation on juvenile Chinook Salmon by Northern Pikeminnow, but presumably, juvenile steelhead and Sockeye Salmon are also prey items in Deadwater Slough, including hatchery releases. Therefore, adult returns of steelhead and Sockeye Salmon are affected as well, including for hatchery populations which provide recreational fishing opportunities. Consequently, reducing predation mortality at Deadwater Slough could potentially benefit multiple upriver natural and hatchery populations, including multiple ESA-listed species. Because the Deadwater Slough is part of the migratory pathway for multiple species and populations of emigrating salmonids, the impact of that benefit could be higher than tributary rehabilitation actions which typically benefit a single population. Moreover, the deepened, slack water conditions that favor Northern Pikeminnow at Deadwater Slough are a result of the failure of a manmade mining reservoir dam. Taken together, the Deadwater Slough is a favorable candidate for management or restoration actions to benefit local Chinook Salmon populations.

We envision two potential management actions at reducing predation at Deadwater Slough: 1) removing the Dump Creek alluvial fan and 2) a local Northern Pikeminnow bounty program to encourage harvest in Deadwater Slough aimed at reducing the predator population size. Both actions have evident pros and cons. Removing the Dump Creek alluvial fan has the benefit of restoring the Salmon River to a more natural flow (e.g., increased water velocity, decreased depth) and bathymetry (e.g., heterogeneous substrate) that likely existed in the reach prior to the failure of the mining reservoir dam in Dump Creek. Restoring flow would presumably reduce Northern Pikeminnow densities to levels similar to upstream and downstream reaches where recent studies identified higher survival and transition probabilities than those at Deadwater Slough (@Axel2015; @Ackerman2018; @Porter2019). However, we do not know the feasibility of removing or alleviating the Dump Creek alluvial fan, and if it could be removed or altered, future disturbance events could potentially move additional loose sediment in the Dump Creek drainage, thereby creating a new alluvial fan. Additionally, restoring the flow would likely diminish the value of the reach as an important bird watching and nesting area.

Northern Pikeminnow bounty programs have proven successful at reducing population sizes and impacts on emigrating salmonids elsewhere (*citation*). A local bounty program in the Salmon River has the potential to boost the local economy (to an unknown degree) which has been negatively impacted by low returns of Chinook Salmon, steelhead, and Sockeye Salmon in recent decades. Many local businesses rely on license and gear sales for local sport fisheries, but low salmon and steelhead returns have eliminated or reduced the duration of fisheries in some years, and a local Northern Pikeminnow bounty program could allow the local economy to recoup some of those lost revenues. On the downside, bounty programs require continual effort and harvest to suppress the population size and so would require years of continued funding and local buy-in to remain effective. If funding or effort was alleviated, Northern Pikeminnow would likely recover to a higher population size without additional alterations to the habitat. 
<!--
What about adding cover? It could potentially provide refugia for juvenile salmonids, but could make recreational angling more difficult

Is there anything in the literature that discusses habitat preferences for pike minnow? If we speed up velocities and add some cover potentially pike minnow predation success will be lowered. Will these fish just move elsewhere? 
-->

@@

The failure of a mining reservoir dam in the late 19th century created an alluvial fan at the confluence of the Salmon River and Dump Creek which has backed up the Salmon River and created an almost 2-kilometer section of unnaturally slow and deep river, resulting in favorable conditions for Northern Pikeminnow. We estimate that greater than 10,000 Northern Pikeminnow occupy Deadwater Slough with densities greater than reported elsewhere. Additionally, the slow water velocity and lack of cover have created conditions that favor predation on juvenile salmonids emigrating through Deadwater Slough including Chinook Salmon i.e., juvenile Chinook Salmon have slower movement rates and little refugia to avoid predation. We estimate that predation from Northern Pikeminnow in Deadwater Slough may reduce adult returns by as much as 146 or 26% of recent returns to the Upper Salmon River MPG. Therefore, predation in Deadwater Slough likely has a consequential impact on ESA-listed Chinook Salmon populations in the Upper Salmon MPG.

>>>>

All mark-recapture abundance estimators suggest a large population of Northern Pikeminnow occupy Deadwater Slough. The mean population size of Northern Pikeminnow in the Deadwater Slough was estimated to be ***value*** during the fall DSR emigration and ***value*** during the spring NRR emigration. Those estimates translate to densities (fish/km) of Northern Pikeminnow that are ***value*** and ***value*** times greater than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately ***value*** and ***value*** times greater than projections for the lower Columbia River [@Beamesderfer1996]. @Beamesderfer1991 acknowledge that their methods were unsuitable for sampling offshore, noting that water velocity, depth, irregular bottom contours, and barge traffic made sampling offshore ineffective; therefore it is possible that their estimates may have been low. Conversely, at Deadwater Slough the maximum depth was approximately 6 m with a relatively homogenous bottom contour, making angling an effective method throughout the entire reach. Given the size (volume) of the Salmon River at the slough relative to the Columbia River, the estimated Northern Pikeminnow densities for Deadwater Slough are remarkable. Further, our estimate densities of Northern Pikeminnow suggest that slow water reaches outside of the reservoir complexes on the Snake and Columbia rivers may support exceptionally high predator densities, consistent with findings of @Harnish2014, @Gray2001, and @Zimmerman1999. Given the dearth of current data available in the Columbia River basin, it is unclear how Northern Pikeminnow abundance estimates in this study compare to elsewhere in the Columbia River Basin; contemporary information on piscine predator abundances may be needed [@Widener2021].

<!--
The mean population size of Northern Pikeminnow in the Deadwater Slough during the fall season was estimated to be `r prettyNum(unique(round(impact_df$N[impact_df$season == "Fall"])), big.mark = ",")` and during the spring 2021 to be `r prettyNum(unique(round(impact_df$N[impact_df$season == "Spring"])), big.mark = ",")`. Those estimates translate to linear densities of `r prettyNum(round(unique(impact_df$N[impact_df$season == "Fall"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(unique(impact_df$N[impact_df$season == "Spring"]) / ddw_rkms), big.mark = ",")` 
which is `r round(unique(impact_df$N[impact_df$season == "Fall"]) / ddw_rkms / john_day_dens)` and `r round(unique(impact_df$N[impact_df$season == "Spring"]) / ddw_rkms / john_day_dens)` times greater, respectively, than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately `r as.character(english(round(unique(impact_df$N[impact_df$season == "Fall"]) / ddw_rkms / low_columbia_fish_km)))` and `r as.character(english(round(unique(impact_df$N[impact_df$season == "Spring"]) / ddw_rkms / low_columbia_fish_km)))` times greater than projections for the lower Columbia River [@Beamesderfer1996]. Although estimated densities for Deadwater Slough are high given the relative size (volume) of the Salmon River at the slough relative to the Columbia River, high CPUE combined with low recapture rates (one in every `r round(nrow(cmr_df) / sum(cmr_df$capture_type == "Recapture", na.rm = T))` captured fish were recaptures) suggest this area may support exceptionally high predator densities. Further, @Beamesderfer1991 acknowledge that their methods were unsuitable for sampling offshore, noting that water velocity, depth, irregular bottom contours, and barge traffic made sampling offshore ineffective. Conversely, at Deadwater Slough the maximum depth was approximately 10-15 m with a relatively homogenous bottom contour, making angling an effective method throughout the entire reach. Our estimates of Northern Pikeminnow areal density are also higher than those reported elsewhere in the Columbia River basin. We estimate the density of Northern Pikeminnow in Deadwater Slough, as fish per hectare, to be `r prettyNum(round(unique(impact_df$N[impact_df$season == "Fall"]) / ddw_hectares), big.mark = ",")` for fall and `r prettyNum(round(unique(impact_df$N[impact_df$season == "Spring"]) / ddw_hectares), big.mark = ",")` for spring. For comparison, the only other available estimates of areal density, that the authors are aware of, are 4.4 fish per hectare for John Day Reservoir [@Beamesderfer1991] and 15 fish per hectare in Lake Washington [@Bartoo1977]. Given the dearth of current data available in the Columbia River basin, it is unclear how Northern Pikeminnow abundance estimates in this study compare to elsewhere in the Columbia River Basin. @Widener2021 has identified a need for more current information on piscine predator populations in the Columbia River.
-->

Several assumptions in our estimators may have influenced the magnitude of our abundance results. First, all models assumed a closed population, however, it is possible that immigration and emigration occurred. Nevertheless, our estimators still provide unbiased estimates of abundance assuming the immigration and emigration rates are equal between marked and unmarked fish. Given the size of the sampling area, and the short duration of our surveys, that assumption is likely met. The assumption of a closed population can also affect results if mortality occurs for some marked fish released back to the population. During our field processing, no mortalities or injury post release was observed; however, the potential for mortality amongst marked and released fish cannot be discounted. Next, our estimates assumed that capture and recapture events are random samples of the population. Angling methods may have a size selection bias, thereby limiting our abundance estimates to a size range susceptible to angling, resulting in a conservative estimate of the total population. We also question the validity of our assumption of equal catchability of individual fish between sampling events. During our study, we recaptured two individual fish during multiple surveys indicating that marked fish were still susceptible to angling.<!--We are also comfortable with the assumptions of no loss of marks or no mistaken or missed marks given that marks were physical (hole punches in fins) and that each fish was examined on a measuring board to collect lengths.--> However, we suspect that marked and released fish may not have mixed thoroughly back into the population prior to recapture events. Although we attempted to release marked fish evenly throughout Deadwater Slough, we anecdotally observed anglers congregating in particular areas of the slough to socialize or exploit the "good" fishing locales. Violations of the assumption of equal catchability between marked and unmarked fish may lead to an overestimate of abundance of Northern Pikeminnow in our study if marked fish have lower catchability. This bias will be proportional to the difference in catchability of marked and unmarked fish, so if marked fish are only 70% as likely to be caught as unmarked fish (due to behavior or spatial distribution) the true abundance is closer to 70% of our abundance estimate.

@@

The spring abundance estimate was smaller than fall abundance estimates. Although our approach assumed equal capture probabilities between the fall and spring surveys, we believe higher spring flows may result in lower capture probability. Therefore, we believe our spring abundance estimate to be conservative.

In addition to the large population, the observed PSD (***value***%) in this study was large compared to observations of 41% below Bonneville Dam and 18% in Bonneville Reservoir [@Winther2020]. This suggests that a larger fraction of Northern Pikeminnow in Deadwater Slough are of a quality size relative to populations reported elsewhere in the Columbia River. Notably, the state catch-and-release record Northern Pikeminnow, measuring ***value*** mm TL, was caught in Deadwater Slough during the fall 2020 survey.

@@

We performed gastric lavage on nearly all Northern Pikeminnow collected during this study. Observed prey species included juvenile Chinook Salmon, species included Redside Shiner *Richardsonius balteatus*, Largescale Sucker *Catostomus macrocheilus*, sculpin *Cottus* spp., and Mountain Whitefish *Prosopium williamsoni*, though decomposition from digestion rendered most fish contents unidentifiable. Although our approach provided useful information on prey species, we have concerns about using angling and gastric lavage to quantify diet composition in Northern Pikeminnow. First, diets of angled fish can differ from those collected by other methods such as electrofishing [@Hodgson1988]. Further, angled fish are more likely to have empty stomachs [@Jurajda2016], perhaps because they are hungry and actively searching for food. Second, unlike other predators such as Smallmouth Bass and Walleye, that are commonly sampled using gastric lavage [@Kamler2001], Northern Pikeminnow are cyprinids and therefore lack a true stomach. Consequently, performing gastric lavage on cyprinid species can be ineffective for accurately quantifying diet composition [@Hartleb1995], or alternately, requires a modification in methodology to flush food items out the vent of the fish [@Wasowicz1994]. Therefore, we consider our diet composition estimates and subsequent consumption estimates to be conservative.

## Bioenergetics

Our approach assumed that all available prey consumed by Northern Pikeminnow have energy densities equal to juvenile Chinook Salmon. Generalizing energy densities using a single species is a common approach [@Petersen1999], however, we recognize that other salmonids including juvenile steelhead and Sockeye Salmon are likely also available prey in Deadwater Slough, especially during the spring migration. Additionally, hatchery smolt releases (Chinook Salmon, steelhead, and Sockeye Salmon) are also prevalent in the mainstem Salmon River during the spring migration. To address the prevalence of other fish in the Northern Pikeminnow's diet, we considered scenarios where DSR and NRR Chinook Salmon represent only 30%, 50%, and 65% of total fish prey consumed. These values may be considered conservative during the emigration periods when Chinook Salmon are abundant and the Northern Pikeminnow diet shifts almost entirely to piscivory, presumably on juvenile Chinook Salmon [@Shively1996; @Poe1991]. This is likely the case during the fall when DSR Chinook Salmon are the dominant prey species available in Deadwater Slough as 1) no (or at least minimal) hatchery releases are present in the river and 2) fall emigrations of steelhead and Sockeye Salmon are less prominent that in Chinook Salmon. During spring when natural-origin steelhead and Sockeye Salmon are also actively emigrating from the Upper Salmon River and hatchery-origin releases of all three species (Chinook Salmon, steelhead, Sockeye Salmon) are present in the river <50% Chinook Salmon in their diet may be more likely. Future work to quantify and identify the juvenile salmonid in Deadwater Slough during their seasonal migration would be useful to validate the diet composition assumptions used in our model.

@@

Although not formally assessed in this study, avian predators are another potential source of mortality for juvenile Chinook Salmon and other ESA-listed salmonids in the Deadwater Slough. The Deadwater Slough is recognized as an important bird watching and nesting area due to the associated upstream braided channels and riparian and backwater habitats [(Deadwater Slough - Audubon Important Bird Areas)](https://www.audubon.org/important-bird-areas/deadwater-slough#). Several piscivorous bird species have been documented using Deadwater Slough including the Common *Mergus merganser* and Hooded *Lophodytes cucullatus* mergansers, the Great Blue Heron *Ardea herodias*, the Double-crested Cormorant *Phalacrocorax auritus*, and the Belted Kingfisher *Megaceryle alcyon* [@eBird2021]. During the initial sampling effort in fall 2019, a two-person crew walked the entire reach including surrounding and upstream areas scanning for passive integrated transponder (PIT) tags that are implanted into thousands of natural- and hatchery-origin salmonids in the Upper Salmon River, annually. During that informal survey, 10 PIT tags were recovered near active bird nests and in an upstream anastomizing reach where herons and eagles are prevalent, suggesting that mortality may have been a result of avian predation. The PIT tag histories in [PTAGIS](https://www.ptagis.org/) indicate these tags were implanted into a combination of juvenile Chinook Salmon (3), Sockeye Salmon (4), and steelhead (3). Two of the Chinook Salmon and one steelhead were natural-origin, the remaining were of hatchery-origin. Avian predation contributes a major component of the total mortality for yearling Chinook Salmon in some locations in the lower Snake and Columbia rivers, particularly at hydroelectric dams and within reservoirs [@Evans2012; -@Evans2016]; however, we did not observe large colonies of piscivorous birds within the study area. Although we did observe Belted Kingfishers, mergansers, and Great Blue Herons at Deadwater Slough; Great Blue Herons were especially prevalent in the anastomizing reach immediately upriver of the slack water reach. Additionally, there is documentation of individual Double-crested Cormorants [@eBird2021] at the Deadwater Slough; however, the site is not within their breeding range, rather, it is part of a migration corridor. Given the current avian species known to occupy Deadwater Slough, it is unlikely that avian predation on juvenile salmonids is comparable to elsewhere in the Columbia River basin with large piscivorous bird colonies. Nevertheless, we hypothesize that the reservoir-like conditions and associated habitat features at the Deadwater Slough may increase the probability of avian predation on juvenile Chinook Salmon and other ESA-listed salmonids from the many piscivorous birds known to use the site. Future studies of predation would benefit from estimates from piscivorous avian predators.

## Potential Impacts to Salmon Populations

Recovery of Chinook Salmon in the Upper Salmon MPG is desired to support local fisheries and economies. At least five of the eight populations must meet criteria set forth by @McElhany2000 and the @ICTRT2007 for the MPG to be considered viable and for recovery of the Snake River Evolutionary Significant Unit. We estimate that ***value*** juvenile Chinook Salmon will be consumed by Northern Pikeminnow, which translates to ***value*** "adult equivalents" which is approximately ***value***% of contemporary returns to the MPG. Considering the diet scenarios used in our study are likely conservative assumptions during the fall DSR emigration, the estimated number of Chinook Salmon consumed is substantial.

Spawner abundance is perhaps the most important metric considered in determining a population's viability. We estimated that consumption of juvenile Chinook Salmon by Northern Pikeminnow in the Deadwater Slough potentially reduces natural-origin returns to upriver populations by ***value*** to ***value*** adults, annually. This is between ***value***% and ***value***% of the total mean adult escapement<!--of 558--> reported for the Upper Salmon MPG in 2017 - 2019 [@Kinzer2020]. In addition to juvenile Chinook Salmon, juvenile steelhead and Sockeye Salmon, including hatchery releases, are also likely prey items for Northern Pikeminnow in Deadwater Slough. Accordingly, we surmise that adult returns of steelhead and Sockeye Salmon are affected by predation, including for hatchery populations which provide recreational fishing opportunities. Consequently, reducing predation mortality at Deadwater Slough could potentially benefit multiple upriver natural and hatchery populations, including ESA-listed species. Because the Deadwater Slough is part of the migratory pathway for multiple species and populations of emigrating salmonids, the positive impact could be greater than individual tributary rehabilitation actions which typically benefit a single population.

@@

The Deadwater Slough is a favorable candidate for management or restoration actions to benefit the local Chinook Salmon populations. Two potential management actions could reduce predation at Deadwater Slough: 1) removing the Dump Creek alluvial fan and 2) a local Northern Pikeminnow bounty program to encourage harvest in Deadwater Slough aimed at reducing the predator population size. Removing the Dump Creek alluvial fan has the benefit of restoring the natural fluvial processes in the Salmon River that likely existed in the reach prior to the failure of the mining reservoir dam in Dump Creek. Restoring natural processes could reduce Northern Pikeminnow densities to levels similar to upstream and downstream reaches where higher survival and transition probabilities have been observed compared to Deadwater Slough [@Axel2015; @Ackerman2018; @Porter2019]. Managers ought also to consider the feasibility and net benefit of restoring fluvial processes relative to potential losses for recreational fishing and bird watching opportunities.

Northern Pikeminnow bounty programs have proven successful at reducing population sizes and impacts on emigrating salmonids elsewhere (@Winther2019). A local Northern Pikeminnow bounty program could provide monetary incentive for sportfisherman, in addition to boosting the local economy from lost revenues. Bounty programs could be conducted year-round or seasonally to coincide with peak juvenile outmigration(s). Each approach would require continual/annual effort and harvest to suppress the population size and as such, appropriate funding would be needed to support the program.

<!--
What about adding cover? It could potentially provide refugia for juvenile salmonids, but could make recreational angling more difficult
-->

@@

The failure of a mining reservoir dam in the late 19th century created an alluvial fan at the confluence of the Salmon River and Dump Creek which has backed up the Salmon River and created favorable conditions for Northern Pikeminnow. We estimate that more than ***value*** Northern Pikeminnow occupy Deadwater Slough, resulting in a remarkable density given the size of Deadwater Slough. Additionally, the slow water velocity and lack of cover have created conditions that favor predation on juvenile salmonids that are rearing in or emigrating through Deadwater Slough, including Chinook Salmon. We estimate that predation from Northern Pikeminnow in Deadwater Slough may reduce adult returns by as much as 146 or 26%, and therefore, has a consequential impact on ESA-listed Chinook Salmon populations in the Upper Salmon MPG.
